---
title: "Comparing data.table reshape to duckdb and polars"
author: "Toby Hocking"
date: "2024-03-10"
categories: [tips, tutorials, developer, benchmarks]
draft: false
---

One element of [the NSF POSE grant for
`data.table`](https://rdatatable-community.github.io/The-Raft/posts/2023-10-15-intro_to_grant-toby_hocking/)
is to create benchmarks which can inform users about when `data.table`
could be more performant than similar software. Two examples of
similar software are duckdb and polars, which each provide in-memory
database operations. This post explores the differences in
computational requirements, and in functionality, for data reshaping
operations. 

# Terminology and functions in R, `data.table`, and SQL

Data reshaping means changing the shape of the data, in order to get
it into a more appropriate format, for learning/plotting/etc. In R we
use the terms "wide" (many columns, few rows) and "long" (few columns,
many rows) to describe the different data shapes (and these terms come
from `?stats::reshape`), whereas in SQL we use the terms "pivoted" and
"unpivoted" to describe these two table types.

| R table type | SQL table type | rows | columns |
|--------------|----------------|------|---------|
| long         | unpivoted      | many | few     |
| wide         | pivoted        | few  | many    |

For the wide-to-long reshape operation, `data.table` has `melt()` and
SQL has `UNPIVOT`; for the long-to-wide reshape operation,
`data.table` has `dcast()` and SQL has `PIVOT`.

| Reshape operation | `data.table` function | SQL function |
|-------------------|-----------------------|--------------|
| Wide-to-long      | `melt`                | `UNPIVOT`    |
| Long-to-wide      | `dcast`               | `PIVOT`      |

# Wide-to-long operations

We begin with a discussion of wide-to-long reshape operations, also
known as unpivot in SQL.

## Wide-to-long data reshape (unpivot) using `data.table::melt`

Wide-to-long reshape is often necessary before plotting. It is
perhaps best explained using a simple example. Here we consider the
iris data, which has four numeric columns:

```{r}
library(data.table)
(iris.wide <- data.table(iris))
```

What if we wanted to make a facetted histogram of the numeric iris
data columns, with one panel/facet for each column? With ggplots we
would use `geom_histogram(aes(numeric_variable))`, where
`numeric_variable` would be the column name of a data table containing
all of the numbers that we want to show in the histogram. To construct
that table, we would have to first reshape to "long" (or unpivoted)
format.  To easily understand what the reshape operation does, we show
a subset of the data (first and last rows) below:

```{r}
(two.iris.wide <- iris.wide[c(1,.N)])
```

Note the table above has 8 numbers, arranged into a table of 2 rows
and 4 columns. To reshape these data to "long" (or unpivoted) format,
we can use `data.table::melt`, as in the code below.

```{r}
melt(two.iris.wide, measure.vars=measure(part, dim, sep="."))
```

Note the table above has the same 8 numbers, but arranged into 1
column in a table with 8 rows, which is the desired input for ggplots.
Also note that the reshaped column names (`Petal.Length`,
`Sepal.Width`, etc) each consist of two components, which become two
different columns in the output: `part` (`Sepal` or `Petal`) and `dim`
(`Length` or `Width`). In the code above, we used `sep="."` to specify
that we want to split all of the iris column names using a dot, and
then reshape all of the columns whose names split into the max number
of items. The corresponding column names of the output are specified
as the arguments of `measure()`, and for more info about this
functionality, please read [its man
page](https://rdatatable.gitlab.io/data.table/reference/measure.html).

Below we do the same reshape with the full iris data set, this time
using a regular expression (instead of the `sep` argument used above),
```{r}
(iris.long <- melt(iris.wide, measure.vars=measure(part, dim, pattern="(.*)[.](.*)")))
```

In the code above, the `pattern` argument is a Perl-compatible regular expression, and columns that match the pattern will be reshaped. The pattern must contain the same number of capture groups (parentheses) as the number of other arguments to melt (part and dim), which are used for output column names. 
After reshaping, we plot the data in a histogram:

```{r dcast-hist}
library(ggplot2)
ggplot()+
  geom_histogram(aes(
    value),
    bins=50,
    data=iris.long)+
  facet_grid(part ~ dim, labeller=label_both)
```

We can see in the plot above that there is a top strip for each `dim`
and a right strip for each `part`, and each facet/panel contains a
histogram of the corresponding subset of data.

## Wide-to-long reshape via unpivot in polars

`polars` is an implementation of data frames in Rust, with bindings in R
and Python. In `polars`, the wide-to-long data reshape operation is
documented on the [man page for
unpivot](https://docs.pola.rs/api/python/stable/reference/dataframe/api/polars.DataFrame.unpivot.html#polars.DataFrame.unpivot), which explains that we must specify `index` and/or `on`  (no support for separator, nor regex). In
our case, we use the code below:

```{r}
(iris.long.polars <- polars::as_polars_df(iris)$unpivot(
  index="Species",
  on=c("Sepal.Length","Petal.Length","Sepal.Width","Petal.Width"),
  variable_name="part.dim",
  value_name="cm"))
```

The output above is analogous to the result from `data.table::melt`,
but with one column named `part.dim` instead of the two columns named
`part` and `dim`, because polars does not support splitting the
reshaped column names into more than one output column. So with
polars, if we wanted separate `part` and `dim` columns, we would have
to specify that in a separate step, after the reshape. Or we could
just use `facet_wrap` instead of `facet_grid`, as in the code below:

```{r polars-hist}
ggplot()+
  geom_histogram(aes(
    cm),
    bins=50,
    data=iris.long.polars)+
  facet_wrap(. ~ part.dim, labeller=label_both)
```

We can see in the plot above that there is a facet for each of the
variables, but only one `part.dim` strip for each, instead of two strips
(`part` and `dim`), as was the case for the previous plot.

### Wide-to-long reshape via UNPIVOT in duckdb

`duckdb` is a column-oriented database implemented in C++, with an R
package that supports a DBI-compliant SQL interface. That means that
we use R functions like `DBI::dbGetQuery` to get results, just like we
would with any other database (Postgres, MySQL, etc). This is
documented in the [duckdb R API](https://duckdb.org/docs/api/r.html)
docs, which explain how to create a database connection, and then copy
data from R to the database, as in the code below,

```{r}
con <- DBI::dbConnect(duckdb::duckdb(), dbdir = ":memory:")
DBI::dbWriteTable(con, "iris_wide", iris)
```

The [duckdb unpivot man
page](https://duckdb.org/docs/sql/statements/unpivot.html) explains
how to do wide-to-long reshape operations, which requires specifying names of columns to reshape (no support for separator, nor regex).  In our case, we use the
code below:

```{r}
iris.long.duckdb <- DBI::dbGetQuery(con, '
UNPIVOT iris_wide
ON "Sepal.Length", "Petal.Length", "Sepal.Width", "Petal.Width" 
INTO NAME part_dim 
VALUE cm')
str(iris.long.duckdb)
```

Above we use `str` to show a brief summary of the structure of the
output, which is a `data.frame` with 600 rows.
With `duckdb`, the output has one column named `part_dim` (dots in column names are not allowed so we use an underscore here instead), because it does not support splitting the reshaped column names into more than one output column. So with duckdb, if we wanted separate `part` and `dim` columns, we would have to specify that in a separate step, after the reshape. 

## Creating `part` and `dim` columns

Both `polars` and `duckdb` are not capable of producing the separate
`part` and `dim` columns during the reshape operation, but we can
always do it as a post-processing step. One way to do that, by
specifying a separator, would be via `data.table::tstrsplit`, as in
the code below:

```{r}
data.table(iris.long.duckdb)[
, c("part","dim") := tstrsplit(part_dim, split="[.]")
][]
```

The code above first converts to `data.table`, then uses the square
brackets to assign new columns. Inside the square brackets, there is a walrus assignment:

* `,` comma because there is no first argument (no subset, use all rows)
* `c("part","dim")` is the left side of the walrus `:=` assignment,
  which specifies the new column names to create.
* on the right side of the walrus, the result of `tstrsplit(part_dim,
  split="[.]")` is used as the value to assign to the new columns
  (`part_dim` is the column to split, and `"[.]"` is the regex to use
  for splitting). 
* Since `tstrsplit` returns a list of two character vectors, there
  will be two new columns.

Finally after the walrus square brackets, we use another empty square
brackets `[]` to enable printing (there is no printing immediately
after assigning new columns using the walrus operator).

Another way of doing that, by specifying a regex, would be via
`nc::capture_first_df` (recently given the `data.table` [Seal of
Approval](https://rdatatable-community.github.io/The-Raft/posts/2024-08-01-seal_of_approval-nc/)),
as in the code below:

```{r}
nc::capture_first_df(iris.long.duckdb, part_dim=list(
  part=".*",
  "[.]",
  dim=".*"))
```

The code above specifies:

* `capture_first_df`, a function for applying capturing regex to columns of a data frame;
* `iris.long.duckdb` is the input data frame, in which there is the `part_dim` column to split;
* `part=".*", "[.]", dim=".*"` makes the capturing regex; R argument names are used to define the new column names, based on the text captured in the corresponding regex (`".*"` means zero or more non-newline characters).

Both results above are data tables with extra cols `part` and `dim`. For visualization, these data tables could be used with either `facet_grid` or `facet_wrap`, similar to the examples above.

## Reshape into multiple columns

Another kind of wide-to-long reshape involves reshaping into multiple columns. 
For example, in the iris data, we may wonder whether sepals are larger than petals (in terms of both length and width). 
To answer that question, we could make a scatterplot of `y=Sepal` versus `x=Petal`, with a facet/panel for each dimension (`Length` and `Width`). 
In the ggplot system, we would need to compute a data table with columns `Sepal`, `Petal`, and `dim`, and we can do that by specifying the `value.name` keyword to `measure()`, as in the code below:

```{r}
(iris.long.parts <- melt(iris.wide, measure.vars=measure(value.name, dim, sep=".")))
```

Again, the `measure()` function in the code above operates by
splitting the input column names using `sep`, which results in two
groups (`Sepal.Width` split into `Sepal` and `Width`, etc) for each of the measured
columns. The `value.name` keyword indicates that each unique value in
the first group (`Sepal` and `Petal`) should be used as the name of an
output column.
This functionality can be very convenient for some data reshaping tasks, but it is neither supported in `polars`, nor in `duckdb`. 
Going back to our original motivating problem, we can make the scatterplot using the code below,

```{r iris-scatter}
ggplot()+
  theme_bw()+
  geom_abline(slope=1, intercept=0, color="grey")+
  geom_point(aes(
    Petal, Sepal),
    data=iris.long.parts)+
  facet_grid(. ~ dim, labeller=label_both)+
  coord_equal()
```

From the plot above, we see that all of the data points (black) are above the y=x line (grey), so we can conclude that sepals are indeed larger than petals, in terms of both length and width.

## Wide-to-long performance comparison

We may also wonder which data reshaping functions work fastest for
large data. To answer that question, we will use `atime`, which is an
R package that allows us to see how much time/memory is required for
computations in R, as a function of data size `N`. In the `setup`
argument of the code below, we repeat the iris data for a certain
number of rows `N`. The code in the other arguments is run for the time/memory measurement, and is very similar to the code presented in previous sections. One difference is that for `data.table` we use `id.vars` instead of `measure()`, to more closely match the arguments provided to the other unpivot functions (for a more fair comparison).

```{r atime-unpivot}
seconds.limit <- 0.1
unpivot.res <- atime::atime(
  N=2^seq(1,50),
  setup={
    (row.id.vec <- 1+(seq(0,N-1) %% nrow(iris)))
    N.df <- iris[row.id.vec,]
    N.dt <- data.table(N.df)
    polars_df <- polars::as_polars_df(N.df)
    duckdb::dbWriteTable(con, "iris_table", N.df, overwrite=TRUE)
  },
  seconds.limit=seconds.limit,
  "duckdb\nUNPIVOT"=DBI::dbGetQuery(con, 'UNPIVOT iris_table ON "Sepal.Length", "Petal.Length", "Sepal.Width", "Petal.Width" INTO NAME part_dim VALUE cm'),
  "polars\nunpivot"=polars_df$unpivot(index="Species", value_name="cm"),
  "data.table\nmelt"=melt(N.dt, id.vars="Species", value.name="cm"))
unpivot.refs <- atime::references_best(unpivot.res)
unpivot.pred <- predict(unpivot.refs)
plot(unpivot.pred)+coord_cartesian(xlim=c(1e1,1e7))
```

In the plot above, the computation time in seconds is plotted as a function of `N`, the number of input rows to reshape. The horizontal reference line is drawn at `r seconds.limit` seconds, and the `N` highlighted corresponds to the throughput given that time limit. When we compare the `N` values shown for the different methods, we see that `data.table` is comparable to `polars` (within 2x), and both are much faster than `duckdb` (about 10x).

Above there are several confounding factors in the comparison, most notably that data must be copied to `duckdb` and `polars` before and after processing. In contrast, `data.table` provides `setDT` and `setDF` functions, which can convert to/from data tables, without copying. So when data originates in R, or needs to come back to R, we should include the copy time for a more fair comparison. Below we run that comparison:

```{r atime-unpivot-copy}
seconds.limit <- 0.1
unpivot.copy.res <- atime::atime(
  N=2^seq(1,50),
  setup={
    (row.id.vec <- 1+(seq(0,N-1) %% nrow(iris)))
    N.df <- iris[row.id.vec,]
  },
  seconds.limit=seconds.limit,
  "duckdb\ncopy+UNPIVOT"={
    duckdb::dbWriteTable(con, "iris_table", N.df, overwrite=TRUE)
    DBI::dbGetQuery(con, 'UNPIVOT iris_table ON "Sepal.Length", "Petal.Length", "Sepal.Width", "Petal.Width" INTO NAME part_dim VALUE cm')
  },
  "polars\ncopy+unpivot"={
    polars_df <- polars::as_polars_df(N.df)
    polars_unpivot <- polars_df$unpivot(index="Species", value_name="cm")
    as.data.frame(polars_unpivot)
  },
  "data.table\nset+melt"=setDF(melt(setDT(N.df), id.vars="Species", value.name="cm")))
unpivot.copy.refs <- atime::references_best(unpivot.copy.res)
unpivot.copy.pred <- predict(unpivot.copy.refs)
plot(unpivot.copy.pred)+coord_cartesian(xlim=c(1e1,1e7))
```

The result above shows that `data.table` is most efficient in terms of
computation time. In this comparison, `data.table` is clearly faster
than `polars` (about 10x), and much faster than `duckdb` (about 100x).

## Wide-to-long summary of functionality

In this section, we showed that `data.table` provides an efficient and feature-rich implementation of wide-to-long data reshaping. 
* `measure()` allows specification of columns to reshape using either
  a separator or a regular expression pattern. In contrast, `duckdb`
  nor `polars` require specifying input column names (no support for
  separator, nor regex), and output column post-processing, which is
  less convenient.
* The `value.name` keyword can be used to reshape into multiple output
  columns, which is required for some kinds of reshape operations (no
  way to do that in `duckdb`/`polars`).
* `setDT` and `setDF` can be used to avoid un-necessary copies with `data.table`. In contrast, `duckdb`/`polars` require copies to/from regular R memory, which can add significant time/memory requirements.
* `data.table` was fastest and most memory efficient in the comparisons we examined (both with and without consideration of copying).

The table below summarizes support for different features in each software package.

| how to specify        | `data.table`     | `polars`        | `duckdb`    |
|-----------------------|------------------|-----------------|-------------|
| function              | `melt`           | `unpivot`       | `UNPIVOT`   |
| reshape cols          | `measure.vars`   | `on`            | `ON`        |
| other cols            | `id.vars`        | `index`         | -           |
| output name (data)    | `value.name`     | `value_name`    | `VALUE`     |
| output name (columns) | `variable.name`  | `variable_name` | `INTO NAME` |
| separator             | `sep`            | -               | -           |
| regex                 | `pattern`        | no              | no          |
| multiple outputs      | `value.name`     | -               | -           |
| avoid copies          | `setDT`, `setDF` | -               | -           |

# Long-to-wide operations

Another kind of reshape operation starts with a long table (many rows, few cols), and creates a wide table (many cols, few rows). I frequently use this operation when comparing results of machine learning algorithms (computing mean/SD over folds, p-values, etc). For examples of those use cases, please read my blog about [Visualizing prediction error](https://tdhock.github.io/blog/2024/viz-pred-err/). 

## Long-to-wide data reshape using `data.table::dcast`

Here we continue with the iris data example. We will present three different reshape operations involving the iris data. The code below adds a column `flower` which contains the row number.

```{r}
iris.wide[, flower := .I][]
```

Then we do a wide-to-long reshape using the code below (same as previous section),

```{r}
(iris.long.i <- melt(iris.wide, measure.vars=measure(part, dim, sep=".")))
```

The table above has an additional column for `flower`, which we use in
the code below on the left side of the formula (used to define output
rows), along with `part + dim` on the right side of the formula (used
to define output columns). The code below can therefore be used to
reshape the data back into their original wide format:

```{r}
dcast(# wide reshape 1
  data=iris.long.i,
  formula=flower + Species ~ part + dim,
  sep=".")
```

We can see that the result above is almost the same as the original
iris data (but with the columns in a different order). Another kind of
reshape involves computing an aggregation function, such as
`mean`. Note in the code below that `.` on the right side of the
formula indicates a single output column.

```{r}
dcast(# wide reshape 2
  data=iris.long.i,
  formula=Species + part + dim ~ .,
  fun.aggregate=mean,
  sep=".")
```

The output above has a row for every unique combination of `Species`,
`part`, and `dim`, and a column (`.`)` for the mean of the
corresponding data.
The more complex reshape below involves multiple aggregations, and multiple value variables.

```{r}
options(width=100)
dcast(# wide reshape 3
  data=iris.long.parts,
  formula=dim ~ Species,
  fun.aggregate=list(mean,sd),
  value.var=c("Sepal","Petal"))
```

The output above includes two rows, and a column for every unique
combination of `value.var` (`Sepal` or `Petal`), of `fun.aggregate`
(`mean` or `sd`), and of `Species` (`setosa`, `versicolor`,
`virginica`).

## Long-to-wide reshape in polars 

`polars` supports long-to-wide reshape via [the `pivot` method](https://docs.pola.rs/api/python/stable/reference/dataframe/api/polars.DataFrame.pivot.html), as in the code below.

```{r}
(polars.wide <- polars::as_polars_df(
  iris.long.i
)$pivot(# wide reshape 1
  on=c("part","dim"),
  index=c("flower","Species"),
  values="value"))
names(polars.wide)
```

The output above is consistent with the results from `data.table::dcast`, and the original iris data, although the names are unusual (with curly braces and double quotes). The next reshape example below shows that we need to create a dummy variable to use as the `on` argument.

```{r}
polars::as_polars_df(
  iris.long.i[, dummy := "."]
)$pivot(# wide reshape 2
  on="dummy", # have to create dummy var for on.
  index=c("Species","part","dim"),
  values="value",
  aggregate_function="mean")
```

The output above is consistent with the results from `data.table::dcast`.
Currently `polars` only supports a single aggregation function, so we can not calculate both `mean` and `sd` at the same time, but we can at least do the `mean` for multiple `values` in the code below:

```{r}
polars::as_polars_df(
  iris.long.parts
)$pivot(# wide reshape 3
  on="Species",
  index="dim",
  values=c("Sepal","Petal"),
  aggregate_function="mean")#multiple agg not supported.
```

Above we see the result only has 6 columns (for `mean`), whereas the analogous result from `data.table::dcast` above had 12 columns (with additionally the `sd`).

## Long-to-wide reshape in duckdb

`duckdb` supports long-to-wide reshape via [the SQL `PIVOT` command](https://duckdb.org/docs/sql/statements/pivot.html), which can be used to recover the original iris data via the command below:

```{r}
duckdb::dbWriteTable(con, "iris_long_i", iris.long.i, overwrite=TRUE)
iris.wide.again.duckdb <- DBI::dbGetQuery(# wide reshape 1
  con, '
PIVOT iris_long_i 
ON part,dim 
USING sum(value) 
GROUP BY flower,Species 
ORDER BY flower')
str(iris.wide.again.duckdb)
```

We can see that the result above is consistent with the previous sections.
The code below uses `mean` as an aggregation function.

```{r}
DBI::dbGetQuery(# wide reshape 2
  con, '
PIVOT iris_long_i 
USING mean(value) 
AS "." 
GROUP BY Species,part,dim')
```

The result above is consistent with previous results.
Finally, we can do multiple aggregations via the code below, which requires enumerating each combination of aggregation function and input column to aggregate.

```{r}
duckdb::dbWriteTable(con, "iris_long_parts", iris.long.parts, overwrite=TRUE)
DBI::dbGetQuery(# wide reshape 3
  con, '
PIVOT iris_long_parts 
ON Species 
USING 
 mean(Sepal) AS Sepal_mean, 
 stddev(Sepal) AS Sepal_sd, 
 mean(Petal) AS Petal_mean, 
 stddev(Petal) AS Petal_sd 
GROUP BY dim')
```

The result above is consistent with the result from
`data.table::dcast`. Because all combinations of aggregation/columns
must be enumerated, the `duckdb` code is a bit more repetitive than
the corresponding `data.table` code (which is more convenient).

## Long-to-wide performance comparison

Below we conduct an `atime` benchmark to measure the computation time of the reshape operation (without controlling for the copy operation).

```{r atime-pivot}
seconds.limit <- 0.1
pivot.res <- atime::atime(
  N=2^seq(1,50),
  setup={
    (row.id.vec <- 1+(seq(0,N-1) %% nrow(iris.long.i)))
    N.dt <- iris.long.i[row.id.vec]
    N.df <- data.frame(N.dt)
    N_polars <- polars::as_polars_df(N.df)
    duckdb::dbWriteTable(con, "iris_long_i", N.df, overwrite=TRUE)
  },
  seconds.limit=seconds.limit,
  "duckdb\nPIVOT"=DBI::dbGetQuery(con, 'PIVOT iris_long_i USING mean(value) AS "." GROUP BY Species,part,dim'),
  "polars\npivot"=N_polars$pivot(on="dummy", index=c("Species","part","dim"), values="value", aggregate_function="mean"),
  "data.table\ndcast"=dcast(N.dt, Species + part + dim ~ ., mean))
pivot.refs <- atime::references_best(pivot.res)
pivot.pred <- predict(pivot.refs)
plot(pivot.pred)+coord_cartesian(xlim=c(1e1,1e7))
```

The result above shows that `data.table::dcast` is about as fast as
the others (bottom facet), although `duckdb` is slightly faster, and `polars` is
slightly slower (less than 2x). 
Below we run a more complex benchmark which also measures computation time for the copy operation (in addition to the reshape).

```{r atime-pivot-copy}
seconds.limit <- 0.1
pivot.copy.res <- atime::atime(
  N=2^seq(1,50),
  setup={
    (row.id.vec <- 1+(seq(0,N-1) %% nrow(iris.long.i)))
    N.df <- data.frame(iris.long.i[row.id.vec])
  },
  seconds.limit=seconds.limit,
  "duckdb\ncopy+PIVOT"={
    duckdb::dbWriteTable(con, "iris_long_i", N.df, overwrite=TRUE)
    DBI::dbGetQuery(con, 'PIVOT iris_long_i USING mean(value) AS "." GROUP BY Species,part,dim')
  },
  "polars\ncopy+pivot"={
    polars_pivot <- polars::as_polars_df(
      N.df
    )$pivot(# wide reshape 2
      on="dummy", # have to create dummy var for on.
      index=c("Species","part","dim"),
      values="value",
      aggregate_function="mean")
    as.data.frame(polars_pivot)
  },
  "data.table\nset+dcast"=setDF(dcast(setDT(N.df), Species + part + dim ~ ., mean)))
pivot.copy.refs <- atime::references_best(pivot.copy.res)
pivot.copy.pred <- predict(pivot.copy.refs)
plot(pivot.copy.pred)+coord_cartesian(xlim=c(1e1,1e7))
```

The result above shows that `data.table` is quite a bit faster than the others (5x or more). 

## Summary of long-to-wide reshaping

In this section, we showed that `data.table` provides an efficient and feature-rich implementation of long-to-wide data reshaping. 
* The formula interface allows specifying a dot (`.`) which is a convenient way to specify output of only one row/column. In contrast, `polars` requires creating a dummy variable to do that.
* The `fun.aggregate` argument may be a list of functions, each of which will be used on each of the `value.var` (a convenient way of specifying all combinations). In contrast, `duckdb` requires specifying each combination separately (more tedious/error-prone), and `polars` only supports one aggregation function (not a list).

| how to specify | `data.table`     | `polars`             | `duckdb`               |
|----------------|------------------|----------------------|------------------------|
| function       | `dcast`          | `pivot`              | `PIVOT`                |
| rows           | LHS of formula   | `index`              | `GROUP BY`             |
| columns        | RHS of formula   | `on`                 | `ON`                   |
| no columns     | dot `.`          | dummy variable       | omit `ON`              |
| values         | `value.var`      | `values`             | `USING`                |
| aggregation    | `aggregate.fun`  | `aggregate_function` | `USING`                |
| multiple agg.  | all combinations | one function         | specified combinations |

# Conclusion

We have compared the reshaping functions in `data.table` to `duckdb`
and `polars`.  Although all three provide similar functionality for
basic operations, we observed that `data.table` has several advantages
for advanced/complex/efficient operations (reshaping columns which
match a regex/separator, reshaping into multiple columns, avoiding
copies, multiple aggregation). We also observed that the speed of
`data.table` functions is comparable, if not faster, than the other
packages.

## Attribution

Parts of this blog post were copied from [my more extensive comparison
blog](https://tdhock.github.io/blog/2024/collapse-reshape/).

## Session info

```{r}
sessionInfo()
```
